The CrypTool Script: \\ Mathematics and Cryptography \\ \vspace{30pt} {\large Background reading\\*[-8pt] for the free elearnig program CrypTool (version 1.3.10)} \vspace{60pt}

### 2005-07-04 script/de/script-de.tex
german umlauts with dvi fix
overfull boxes and other fixes
be comments
no message
be updates
more be updates
cvs id
be anmerkungen
menu appendix preliminary version
be update
*** empty log message ***
Update - Bernhard Esslinger June 2005

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\documentclass[a4paper,11pt]{article}
\usepackage[german]{babel}
\usepackage{hyperref}
% \usepackage[dvips]{graphics}
% \usepackage{graphicx}

@@

@@

%\usepackage{myThanks}
%\usepackage{thumbpdf}

@@

\setlength{\textwidth}{460 pt}
\hoffset-0.7in
\newtheorem{definition}{Definition}
\newtheorem{satz}{Satz}
\newenvironment{Beweis}[1]
  {\textbf{Beweis #1} \\}
  {\hfill$\Box$ \\}

%\pdfinfo{ /Title (CrypTool Skript) /Creator (TeX) /Producer
%(pdfTeX 0.14a) /Author (Deutsch Bank AG) /CreationDate
%(D:20000920201000) }

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\title{ CrypTool Skript \\ Mathematik und Kryptographie }

\author{
(c) Die Autoren, 1998-2002 \\
Frankfurt am Main \\*[8pt]
}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{document}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

\input{chapters/aboutcryptool.tex}

\newpage
\tableofcontents
\addcontentsline{toc}{section}{Inhaltsverzeichnis}
\newpage
\normalsize
\parindent0cm
\parskip4pt

\input{chapters/introduction.tex}
\input{chapters/cryptomethods.tex}
\input{chapters/primes.tex}
\input{chapters/numbertheory.tex}
\input{chapters/moderncryptography.tex}
\input{chapters/digitalsignatures.tex}
\input{chapters/ellipticcurves.tex}

\printindex

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

>>>>

% $Id$
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\documentclass[a4paper,11pt]{article}
% \usepackage[dvips]{graphics}
% \usepackage{graphicx}

@@

%\usepackage{myThanks}
%\usepackage{thumbpdf}
\usepackage{hyperref}
%\usepackage[T1]{fontenc}
%\usepackage{times}
%\usepackage{pslatex}
%\usepackage{ae,aecompl}

@@

@@

\setlength{\textwidth}{460 pt}
\hoffset-0.7in
\newtheorem{definition}{Definition}[section]
\newtheorem{satz}{Satz}[section]
\newenvironment{Beweis}[1]
  {\textbf{Beweis #1} \\}
  {\hfill$\Box$ \\}

%\pdfinfo{ /Title (CrypTool-Skript) /Creator (TeX) /Producer
%(pdfTeX 0.14a) /Author (Deutsch Bank AG) /CreationDate
%(D:20000920201000) }

\usepackage[german]{babel}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\title{\vskip -20 pt 
	{ Das CrypTool-Skript: \\ 
	Mathematik und Kryptographie} \\
        \vspace{30pt}
        {\large Hintergrundmaterial und Zusatzinformationen\\*[-8pt]
         zum freien eLearning-Programm CrypTool (Version 1.3.10)}
        \vspace{60pt}
}

\vspace{32pt}

\author{
(c) Bernhard Esslinger (Mitautor und Herausgeber), 1998-2005 \\
Frankfurt am Main \\*[8pt]
}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{document}

@@

% Vergleichbar zur Nutzung von Texten aus der Wikipedia (=GNU Free Documentation Licence, GFDL)
\vspace{200pt}
Dies ist ein freies Dokument, d.h. Inhalte des Dokuments k"onnen kopiert und verbreitet werden,
auch zu kommerziellen Zwecken. \\
Im Gegenzug sind Autor, Titel und die CrypTool-Webseite zu nennen.
Selbstverst"andlich darf aus dem CrypTool-Skript, genau wie aus jedem anderen Werk auch, zitiert werden.
Diese Lizenzbedingungen sind angelehnt an die GNU-Lizenz f"ur freie Dokumentation.

% be_2005: a) Auf erste Seite nur den Titel ! -> Newpage eingefügt.
%          b) "aboutcryptool.tex" hier und nicht unten vor "introduction.tex" einfügn, damit es VOR dem Inhaltsverzeichnis steht.
\newpage
\input{chapters/aboutcryptool.tex}

\newpage
\tableofcontents
\addcontentsline{toc}{section}{Inhaltsverzeichnis}

\newpage
\normalsize
\parindent 0cm
\parskip 4pt

\input{chapters/introduction.tex}
\input{chapters/cryptomethods.tex}
\input{chapters/paper_and_pencil.tex}
\input{chapters/primes.tex}
\input{chapters/numbertheory.tex}
\input{chapters/moderncryptography.tex}
\input{chapters/digitalsignatures.tex}
\input{chapters/ellipticcurves.tex}

%\appendix
\begin{appendix}
  \newpage
  \input{chapters/menus.tex}
  \input{chapters/authors.tex}
\end{appendix}

%\addcontentsline{toc}{section}{Index} %moved to style.ist
\printindex

@@

### 2005-07-04 script/en/chapters/aboutcryptool.tex
Update - Bernhard Esslinger June 2005

% $Id$
% ............................................................................
%                 TEXT DER 1. SEITE
% ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~

\parskip 4pt
%\vskip +12 pt
{
\noindent In this CrypTool\index{CrypTool} script you will find predominantly
mathematically oriented information on using cryptographic procedures.
The main chapters have been written by various authors
% \hyperlink{appendix-authors}{(authors)}
% in \ref{s:appendix-authors}
% \hypertarget{appendix-authors}{}\label{s:appendix-authors}
and are therefore independent from one another. At the end of most chapters
you will find literature and web links.

You will obtain information about the principles of symmetrical and
asymmetrical {\bf encryption}. A large section of this script is dedicated
to the fascinating topic of {\bf prime numbers}. Using numerous examples,
the {\bf elementary number theory} and {\bf modular arithmetic} are
introduced and applied in an exemplary manner for the {\bf RSA procedure}.
By reading the following chapter you'll gain an insight into the mathematical
ideas behind {\bf modern cryptography}.

A further chapter is devoted to {\bf digital signatures}, 
which are an essential component of e-business applications. 
The last chapter describes {\bf elliptic curves}: they could be used
as an alternative to RSA and in addition are extremely well suited for 
implementation on smartcards.

Whereas the \textit{program} CrypTool\index{CrypTool} teaches you how to use
cryptography in practice, the \textit{script} provides those interested in the
subject with a deeper understanding of the mathematical algorithms
used -- trying to do it in an instructive way.

The authors Bernhard Esslinger, Matthias B\"uger, Bartol Filipovic, 
Henrik Koy, Roger Oyono and J\"org-Cornelius Schneider
would like to take this opportunity to thank their colleagues 
in the company and at the universities of Frankfurt, Gie\ss en, 
Siegen, Karlsruhe and Darmstadt. They are particularly indepted to Dr.~Peer
Wichmann from the Karlsruhe computer science research centre (Forschungszentrum
Informatik, FZI) for his down-to-earth support.

\enlargethispage{12pt}
As at CrypTool\index{CrypTool}, the quality of the script is enhanced
by your suggestions and ideas for improvement. We look forward 
to your feedback.

\vskip +7 pt \noindent
You will find the current version of CrypTool\index{CrypTool} under \newline
  \href{http://www.cryptool.org}{\texttt{http://www.cryptool.org}},~
  \href{http://www.cryptool.com}{\texttt{http://www.cryptool.com}}~ or~
  \href{http://www.CrypTool.de}{\texttt{http://www.cryptool.de}}.

\vskip +7 pt \noindent
The contact people for this free open-source tool are listed 
in the ``readme'' file delivered within the CrypTool\index{CrypTool} package.}

>>>>

% $Id$
% ............................................................................
%                 TEXT OF THE 2nd PAGE
% ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~

% --------------------------------------------------------------------------
\section*{Overview about the Content of the CrypTool Script}  
\addcontentsline{toc}{section}{Overview}

\parskip 4pt
%\vskip +12 pt
{
\noindent In this CrypTool\index{CrypTool} script you will find predominantly
mathematically oriented information on using cryptographic procedures.
The main chapters have been written by various
{\bf authors} (see appendix \ref{s:appendix-authors})
%\hyperlink{appendix-authors}{authors}
and are therefore independent from one another. At the end of most chapters
you will find literature and web links.

The \hyperlink{Kapitel_1}{first chapter} explains the principles of symmetric
and asymmetric {\bf encryption} and describes shortly the current decryption
records of modern symmetric algorithms.

Because of didactical reasons the \hyperlink{Kapitel_PaperandPencil}
{second chapter} gives an exhaustive overview
about {\bf paper and pencil encryption methods}.

Big parts of this script are dedicated to the fascinating topic of 
{\bf prime numbers} (chap. \ref{Label_Kapitel_2}).
%\hyperlink{Kapitel_2}{{\bf prime numbers}}. 
Using numerous examples,
{\bf modular arithmetic} and 
{\bf elementary number theory} (chap. \ref{Chapter_ElementaryNT})
are introduced and applied in an exemplary manner for the {\bf RSA procedure}.

By reading chapter \ref{Chapter_ModernCryptography}
you'll gain an insight into the mathematical ideas and concepts behind 
{\bf modern cryptography}.

A \hyperlink{Chapter_Hashes-and-Digital-Signatures}{further chapter} gives
an overview about the status of attacks against modern {\bf hash algorithms}
and is then shortly devoted to {\bf digital signatures}, 
which are an essential component of e-business applications. 
The \hyperlink{ellcurve}{last chapter} describes {\bf elliptic curves}:
they could be used as an alternative to RSA and in addition are extremely
well suited for implementation on smartcards.

Whereas the \textit{eLearning program} CrypTool\index{CrypTool} motivates and
teaches you how to use cryptography in practice, the \textit{script} provides
those interested in the subject with a deeper understanding of the mathematical
algorithms used -- trying to do it in an instructive way.
If you are alrady a little bit familiar with this field of knowledge you can
gain a fast overview about the functions delivered by CrypTool using the
{\bf menu tree} (see appendix \ref{s:appendix-menutree}).

% Bernhard Esslinger, Matthias B\"uger, Bartol Filipovic, Henrik Koy, 
% Roger Oyono and J\"org Cornelius Schneider
The authors would like to take this opportunity to thank their colleagues 
in the company and at the universities of Frankfurt, Gie\ss en, 
Siegen, Karlsruhe and Darmstadt.

\enlargethispage{12pt}
As with the eLearning program CrypTool\index{CrypTool}, the quality of the 
script is enhanced by your suggestions and ideas for improvement. 
We look forward to your feedback.

\vskip +7 pt \noindent
You will find the current version of CrypTool\index{CrypTool}
at the web sites \newline
  \href{http://www.cryptool.org}{\texttt{http://www.cryptool.org}},~
  \href{http://www.cryptool.com}{\texttt{http://www.cryptool.com}}~ or~
  \href{http://www.CrypTool.de}{\texttt{http://www.cryptool.de}}.

\vskip +7 pt \noindent
The contact people for this free open-source program are listed at the 
web site and in the ``readme'' file delivered within the 
CrypTool\index{CrypTool} package.}

### 2005-07-04 script/en/chapters/authors.tex
Update - Bernhard Esslinger June 2005

\section{Authors of this script}
This appendix lists the authors of this document. Please refer to the top of the individual
chapters for their contribution.

\begin{description}
\item[Bernhard Esslinger,] initiator of CrypTool, head IT security at Deutsche
Bank and lecturer on IT security at the University of Siegen.

\item[Matthias B\"uger,] contribution to the chapter ``Elliptic Curves'',
research analyst at Deutsche Bank and responsible for Deutsche Bank's
crypto standard.

\item[Bartol Filipovic,] original author of the CrypTool elliptic curve
implementation and the corresponding chapter in this script.

\item[Henrik Koy, ] main developer and co-ordinator of CrypTool development
since version 1.3; script reviewer and \TeX\ guru; project leader IT at
Deutsche Bank.

\item[Roger Oyono, ] implementer of the CrypTool factorization dialogue and
original author of chapter ``The Mathematical Ideas behind Modern
Cryptography''.

\item[J\"org Cornelius Schneider,] design and support of CrypTool; crypto
enthusiast and project leader IT at Deutsche Bank.

@@

>>>>

% ++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++
\subsection{Authors of the CrypTool Script\index{CrypTool}}
\hypertarget{appendix-authors}{}\label{s:appendix-authors}

This appendix lists the authors\index{Authors} of this document.\\
Please refer to the top of each individual chapter for their contribution.

\begin{description}

\item[Bernhard Esslinger,] \mbox{}\\
initiator of the CrypTool project, main author of this script, head IT security
at Deutsche Bank and lecturer on IT security at the University of Siegen.
E-mail: besslinger@web.de.

\item[Matthias B\"uger,] \mbox{}\\ 
contribution to the chapter ``Elliptic Curves'', research analyst at Deutsche Bank.

\item[Bartol Filipovic,] \mbox{}\\
original author of the CrypTool elliptic curve
implementation and the corresponding chapter in this script.

\item[Henrik Koy, ] \mbox{}\\
main developer and co-ordinator of CrypTool development
since version 1.3; script reviewer and \TeX\ guru; cryptographer 
and project leader IT at Deutsche Bank.

\item[Roger Oyono, ] \mbox{}\\
implementer of the CrypTool factorization dialogue and
original author of chapter ``The Mathematical Ideas behind Modern
Cryptography''.

\item[J\"org Cornelius Schneider,] \mbox{}\\
design and support of CrypTool; crypto
enthusiast and senior project leader IT at Deutsche Bank.

\item[Christine St\"otzel,] \mbox{}\\
Master of Business and Computer Science at the University of Siegen.

@@

% ++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++
\newpage
\subsection{Bibliography of Movies and Literature with Relation to Cryptography}
\hypertarget{appendix-movies}{}
\label{s:appendix-movies}
% {\bf movies and literature related to cryptography}(see appendix \ref{s:appendix-movies})
\index{Movies}
\index{Literature}

Cryptographic applications -- classical as well as modern ones -- have been
used in literature and movies. In some media they are only mentioned and
are a pure addmixture; in others they play a primary role and are explained
in detail; and sometimes the the purpose of the story, which forms the framework,
is primarily to transport this knowledge and achieve better motivation.

Here is the beginning of an overview.

\begin{description}

\item[\textrm{[Doyle1905]}] \index{Doyle 1905}
    Arthur Conan Doyle \index{Doyle, Sir Arthur Conan}, \\
    {\em The Adventure of the Dancing Men}, 1905. \\
    In this Sherlock-Holmes short story (first published in 1903 in the
    ``Strand Magazine'', and then in 1905 in the collection 
    ``The Return of Sherlock Holmes'' the first time in book-form)
    Sherlock Holmes has to solve a cipher which at first glance looks
    like a harmless kid's picture. \\
    But it proves to be the
    monoalphabetic substitution cipher of the criminal Abe Slaney.
    Sherlock Holmes solves the riddle using frequent analysis.

\item[\textrm{[Sayer1932]}] \index{Sayer 1932}
    Dorothy L. Sayer, \\
    {\em Have his carcase}, Harper / Victor Gollancz Ltd., 1932. \\
    In this novel the writer Harriet Vane finds a dead body at the beach.
    The police believe the death is suicide.
    Harriet Vane and the elegant amateur sleuth Lord Peter Wimsey together
    clear of the repellent murder in this second of Sayers's famous
    Harriet Vane mystery series. \\
    This requires to solve a cryptogram. Surprisingly the novel not only
    describes the Playfair cipher in detail, but also the cryptanalysis
    of this cipher.

\item[\textrm{[Seed1990]}] \index{Seed 1990}
    Directed by Paul Seed, \\
    {\em House of Cards}, 1990. \\
    In this movie Ruth tries to solve the secret, which made her daughter
    fall silent. Here two young people suffering from autism communicate
    via 5- and 6-digit primes.
    After more than 1 hour the movie contains the following undecrypted
    two series of primes:
    \begin{center}
    $21,383; \;\;176,081; \;\;18,199; \;\;113,933; \;\;150,377; \;\;304,523; \;\;113,933$ \\
    $193,877; \;\;737,683; \;\;117,881; \;\;193,877$
    \end{center}
    % where autistic children communicate via primes

\item[\textrm{[Robinson1992]}] \index{Robinson 1992}
    Directed by Phil Alden Robinson, \\
    {\em Sneakers}, Universal Pictures Film, 1992. \\
    In this movie the ``sneakers'', computer experts under their boss Martin
    Bishop, try to get back the deciphering box SETEC from the ``bad guys''.
    SETEC, invented by a genius mathematician before he was killed, allows to
    decrypt all codes from any nation. \\
    The code is not described in any way.

\item[\textrm{[Elsner1999]}] \index{Elsner 1999}
    Dr.~C.~Elsner, \\
    {\em The Dialogue of the Sisters}, c't, 1999. \\
    In this short story, which is included in the CrypTool package
    \index{CrypTool} as PDF file, the sisters confidentially communicate
    using a variant of RSA.

\item[\textrm{[Stephenson1999]}] \index{Stephenson 1999}
    Neal Stephenson, \\
    {\em Cryptonomicon}, Harper, 1999. \\
    This thick novel deals with cryptography both in WW2 and today.
    The two heros from the 40th are the excellent mathematician and
    cryptanalyst Lawrence Waterhouse, and the overeager and addicted
    US marine Bobby Shaftoe. 
    They both are members of the special allied unit 2702, which tries,
    to hack the enemy's communication codes and at the same time to
    hide the own existance. \\
    This secretiveness also happens in the present plot, where the 
    grandchidren of the war heros -- the dedicated programmer  
    Randy Waterhouse and the beautiful Amy Shaftoe -- team up. \\
    Cryptonomicon is notably heavy for non-technical readers in parts.
    Several pages are spent explaining in detail some of the concepts
    behind cryptography.
    Stephenson added a detailled description of the Solitaire cipher
    (called "Pontifex" in the book), a paper and pencil encryption
    algorithm\index{Paper- and pencil methods} developed by Bruce Schneier.
    The used modern algorithm is not revealed.

\item[\textrm{[Elsner2001]}] \index{Elsner 2001}
    Dr.~C.~Elsner, \\
    {\em The Chinese Labyrinth}, c't, 2001. \\
    In this short story, which is included in the CrypTool package
    \index{CrypTool} as PDF file, Marco Polo has to solve problems from
    number theory within a competition to become a major consultant of
    the Great Khan.

\item[\textrm{[Colfer2001]}] \index{Colfer 2001}
    Eoin Colfer, \\
    {\em Artemis Fowl}, Viking, 2001. \\
    In this book for young people the 12 year old Artemis, a genius thief, gets
    a copy of the top secret ``Book of the Elfs''. After he decrypted it with
    his computer, he finds out things, men never should have known. \\
    The used code is not described in detail or revealed.

\item[\textrm{[Howard2001]}] \index{Howard 2001}
    Ron Howard, \\
    {\em A Beautiful Mind}, 2001. \\
    This is the film version of Sylvia Nasar's biography of the game theorist
    John Nash. 
    After the brilliant but asocial mathematician accepts secret work in 
    cryptography, his life takes a turn to the nightmarish.
    His irresistible urge to solve problems become a danger for himself and
    his family. Nash is -- within his belief -- ein most important hacker working
    for the government.\\
    Details of his way analysing code are not described in any way.

\item[\textrm{[Apted2001]}] \index{Apted 2001}
    Directed by Michael Apted, \\
    {\em Enigma}, 2001. \\
    This is the film version of Robert Harris' ``historical fiction'' 
    {\em Enigma} (Hutchinson, London, 1995) about the World War II 
    codebreaking work at Bletchley Park in early 1943, when the actual
    inventer of the analysis Alan Turing (after Polish pre-work) already 
    was in the US.
    So the fictional mathematician Tom Jericho is the lead character
    in this spy-thriller.\\
    Details of his way analysing the code are not described.

\item[\textrm{[Kippenhahn2002]}] \index{Kippenhahn 2002}
    Rudolf Kippenhahn, \\
    {\em Top secret! -- How to encrypt messages and to hack codes (original
    title: Streng geheim! -- Wie man Botschaften verschl"usselt und 
    Zahlencodes knackt)}, rororo, 2002. \\
    In this novel a grandpa, an expert for secret writings teaches his
    four grandchildren and their friends, how to encrypt messages which
    nobody should read. Because there is someone who hacks their secrets,
    the grandpa has to teach them more and more complicated methods. \\
    Within this story, which forms the framework, the most important classic
    encryption methods and its analysis are explained in manner, exciting
    and appropriate for children.

\item[\textrm{[Isau2003]}] \index{Isau 2003}
    Ralf Isau, \\
    {\em The Museum of the stolen memories (original title: Das Museum
    der gestohlenen Erinnerungen)}, Thienemann-Verlag, 2003. \\
    In this novel the last part of the oracle can only be
    solved with the joined help of the computer community.

\item[\textrm{[Brown2003]}] \index{Brown 2003}
    Dan Brown, \\
    {\em The Da Vinci Code}, Doubleday, 2003. \\
    The director of the Louvre is found murderered in his museum in
    front of a picture of Leonardo da Vinci. And the symbol researcher
    Robert Langdon is involved in a conspiracy.
    The plot mentions different classic codes (substitution like
    Caesar or Vigenere, as well as transposition and number codes).
    Also there are hints about Schneier and the sunflower.
    The second part of the book contains a lot of theologic considerations.\\
    This book has become one of the most widely read books of all time.

### 2005-07-04 script/en/chapters/cryptomethods.tex
removed nbsps
Update - Bernhard Esslinger June 2005

\newpage
\section{Encryption Procedures}
\hypertarget{Kapitel_1}{}
(Bernhard Esslinger, May 1999, Updates Dec. 2001, Feb. 2003)

@@

% --------------------------------------------------------------------------
\subsection[Symmetric encryption]
{Symmetric encryption\footnotemark}
  \footnotetext{%
    With CrypTool\index{CrypTool} v1.3 you can execute the following modern
    symmetric encryption algorithms 
    (using the menu {\bf Crypt \textbackslash{} Symmetric}): \\
    IDEA, RC2, RC4, DES (ECB), DES (CBC), Triple-DES (ECB), Triple-DES (CBC),
    MARS (AES candidate), RC6 (AES candidate), Serpent (AES candidate), 
    Twofish (AES candidate), Rijndael (official AES algorithm).
  }

@@

All classical methods are of this type. Examples can be found within 
CrypTool or in \cite{Nichols1996}. Now we want to consider more modern 
mechanisms.

@@

Brute-force (exhaustive search, trial-and-error) means to completely examine all keys of the key space: so no special analysis methods have to be used. Instead, the cipher text is decrypted with all possible keys and for each resulting text it is checked, whether this is a meaningful clear text. A key length of 64 bit means at most $2^{64}$ = 18,446,744,073,709,551,616 or about 18 trillion (GB) / 18 quintillion (US)  keys to check\index{CrypTool}%
\footnote{%
    With CrypTool\index{CrypTool} v1.3 you can also try brute-force attacks
    of modern symmetric algorithms 
    (using the menu {\bf Analysis \textbackslash{} Ciphertext-Only}): 
    to achieve a result in an appropriate time with a single PC you should 
    mark not more than 20 bit of the key as unknown.
}.

@@

% --------------------------------------------------------------------------
\subsection[Asymmetric encryption]
{Asymmetric encryption\footnotemark}
  \footnotetext{%
    With CrypTool\index{CrypTool} v1.3 you can execute RSA encryption
    and decryption (using the menu {\bf Crypt \textbackslash{} Asymmetric}).
  }

@@

The most well-known asymmetric procedure is the \index{RSA} 
RSA algorithm\index{CrypTool}%
\footnote{%
The RSA algorithm is extensively described in chapter \ref{rsabeweis} and later
within this script. 
The RSA cryptosystem can be executed in many variations with 
CrypTool\index{CrypTool} v1.3 (using the menu 
{\bf Individual Procedures \textbackslash{} RSA Demonstration \textbackslash{}
RSA Cryptosystem}). 
The topical research results concerning RSA are described 
in chapter \ref{SecurityRSA}.
}%
,
named after its developers Ronald \index{Rivest Ronald} Rivest, Adi
\index{Shamir Adi} Shamir and Leonard \index{Adleman Leonard} Adleman. The RSA algorithm
was published in 1978. The concept of asymmetric encryption was first
introduced by Whitfield Diffie \index{Diffie Whitfield}  and Martin
\index{Hellman Martin} Hellman in 1976. Today, the ElGamal \index{ElGamal}
procedures also play a decisive role, particularly the \index{Schnorr C.P.} Schnorr
variant in the \index{DSA} DSA (Digital \index{Signature!digital}Signature
Algorithm).

% --------------------------------------------------------------------------
% \newpage
\subsection[Hybrid procedures]
{Hybrid procedures\footnotemark}
\footnotetext{%
Within CrypTool\index{CrypTool} v1.3 you can get a visualization of this
technique using the menu {\bf Crypt \textbackslash{} Hybrid Demonstration}: 
this dialogue shows the single steps and its dependencies with concrete
numbers.
}\index{Hybrid procedure}

@@

In this case the data is encrypted using symmetric procedures: the key is a
session key\index{Session key} generated by the sender
randomly\footnote{%
An important part of cryptographically secure techniques is to generate 
random numbers. Within CrypTool\index{CrypTool} v1.3 you can check out
different random number generators using the menu {\bf Indiv. Procedures 
\textbackslash{} Generate Random Numbers}. 
Using the menu {\bf Analysis \textbackslash{} Random Tests} you can apply
different test methods for random data to binary documents. \\
Up to now CrypTool has concentrated on cryptographically strong 
pseudo random number generators. Only the integrated Secude\index{Secude}
generator involves a "pure" random source. 
}\index{Random}
that is only used for this message.
This session key is then encrypted using the asymmetric procedure and
transmitted to the recipient together with the message. Recipients can determine
the session key using their secret keys and then use the session key to encrypt
the message. In this way, we can benefit from the easy key management
\index{Key management} of asymmetric procedures and encrypt large quantities of data
quickly and efficiently using symmetric procedures.

@@

\bibitem[Nichols1996]{Nichols1996} \index{Nichols 1996} 
    Randall K. Nichols, \\
    {\em Classical Cryptography Course, Volume 1 and 2}, \\
    Aegean Park Press 1996.

>>>>

\newpage
\section{Encryption Procedures}
\hypertarget{Kapitel_1}{}
(Bernhard Esslinger, May 1999, Updates Dec. 2001, Feb. 2003, June 2005)

@@

% --------------------------------------------------------------------------
\subsection[Symmetric encryption]
{Symmetric encryption\footnotemark}
  \footnotetext{%
    With CrypTool\index{CrypTool} you can execute the following modern
    symmetric encryption algorithms 
    (using the menu path {\bf Crypt \textbackslash{} Symmetric (modern)}): \\
    IDEA, RC2, RC4, DES (ECB), DES (CBC), Triple-DES (ECB), Triple-DES (CBC),
    MARS (AES candidate), RC6 (AES candidate), Serpent (AES candidate), 
    Twofish (AES candidate), Rijndael (official AES algorithm).
  }

@@

All classical methods are of this type. Examples can be found within the 
program CrypTool, in chapter  \ref{Kapitel_PaperandPencil} 
(Paper and Pencil Encryption Methods) of this script or in \cite{Nichols1996}. 
Now we want to consider more modern mechanisms.

@@

Brute-force (exhaustive search, trial-and-error) means to completely examine all keys of the key space: so no special analysis methods have to be used. Instead, the cipher text is decrypted with all possible keys and for each resulting text it is checked, whether this is a meaningful clear text. A key length of 64 bit means at most $2^{64}$ = 18,446,744,073,709,551,616 or about 18 trillion (GB) / 18 quintillion (US)  keys to check\index{CrypTool}%
\footnote{%
    With CrypTool\index{CrypTool} you can also try brute-force attacks
    of modern symmetric algorithms (using the menu path
    {\bf Analysis \textbackslash{} Symmetric Encryption (modern)}): here
    the weakest knowledge of an attacker is assumed, he performs a 
    ciphertext-only attack.
    To achieve a result in an appropriate time with a single PC you should 
    mark not more than 20 bit of the key as unknown.
}.

@@

% --------------------------------------------------------------------------
\subsection[Asymmetric encryption]
{Asymmetric encryption\footnotemark}
  \footnotetext{%
    With CrypTool\index{CrypTool} you can execute RSA encryption
    and decryption (using the menu path {\bf Crypt \textbackslash{} Asymmetric}).
    In both cases you must select a RSA key pair. Only in the case of decryption
    the secret RSA key is necessary: so here you are asked to enter the PIN. 
  }

@@

The most well-known asymmetric procedure is the \index{RSA} 
RSA algorithm\index{CrypTool}%
\footnote{%
The RSA algorithm is extensively described in chapter \ref{rsabeweis} and later
within this script. 
The RSA cryptosystem can be executed in many variations with 
CrypTool\index{CrypTool} (using the menu path
{\bf Individual Procedures \textbackslash{} RSA Cryptosystem \textbackslash{}
RSA Demonstration}). 
The topical research results concerning RSA are described 
in chapter \ref{SecurityRSA}.
}%
,
named after its developers Ronald \index{Rivest, Ronald} Rivest, Adi
\index{Shamir, Adi} Shamir and Leonard \index{Adleman, Leonard} Adleman. The RSA algorithm
was published in 1978. The concept of asymmetric encryption was first
introduced by Whitfield Diffie \index{Diffie, Whitfield}  and Martin
\index{Hellman, Martin} Hellman in 1976. Today, the ElGamal \index{ElGamal, Tahir}
procedures also play a decisive role, particularly the \index{Schnorr, C.P.} Schnorr
variant in the \index{DSA} DSA (Digital \index{Signature!digital}Signature
Algorithm).

% --------------------------------------------------------------------------
% \newpage
\subsection[Hybrid procedures]
{Hybrid procedures\footnotemark}
\footnotetext{%
Within CrypTool\index{CrypTool} you can get a visualization of this
technique using the menu path {\bf Crypt \textbackslash{} Hybrid}: 
this dialogue shows the single steps and its dependencies with concrete
numbers. Here the asymmetric algorithm RSA and the symmetric algorithm AES are used.
}\index{Hybrid procedure}

@@

In this case the data is encrypted using symmetric procedures: the key is a
session key\index{Session key} generated by the sender
randomly\footnote{%
An important part of cryptographically secure techniques is to generate 
random numbers. Within CrypTool\index{CrypTool} you can check out
different random number generators using the menu path
{\bf Indiv. Procedures \textbackslash{} Generate Random Numbers}. 
Using the menu path {\bf Analysis \textbackslash{} Analyse Randomness}
you can apply different test methods for random data to binary documents. \\
Up to now CrypTool has concentrated on cryptographically strong 
pseudo random number generators. Only the integrated Secude\index{SECUDE IT Security}
generator involves a "pure" random source. 
}\index{Random}
that is only used for this message.
This session key is then encrypted using the asymmetric procedure and
transmitted to the recipient together with the message. Recipients can determine
the session key using their secret keys and then use the session key to encrypt
the message. In this way, we can benefit from the easy key management
\index{Key management} of asymmetric procedures and encrypt large quantities of data
quickly and efficiently using symmetric procedures.

@@

\bibitem[Nichols1996]{Nichols1996} \index{Nichols 1996} 
    Randall K. Nichols, \\
    {\em Classical Cryptography Course, Volume 1 and 2}, \\
    Aegean Park Press 1996;
    or in 12 lessons online at \\
    \href{http://www.fortunecity.com/skyscraper/coding/379/lesson1.htm}
    {\texttt{http://www.fortunecity.com/skyscraper/coding/379/lesson1.htm}}
